### 2025-07-19 evaluating-prediction-accuracy.qmd
20250718 - confusion matrix

Predictions can come in different types.
Some predictions involve categorical data, whereas other predictions involve continuous data.
When dealing with a dichotomous ([nominal data](#sec-nominal) that are binary) [predictor](#sec-correlationalStudy) and [outcome](#sec-correlationalStudy) variable (or continuous data that have been dichotomized using a cutoff), we can evaluate predictions using a 2x2 table known as a confusion matrix (see INSERT), or with logistic regression models.
When dealing with a continuous [outcome variable](#sec-correlationalStudy) (e.g., [ordinal](#sec-ordinal), [interval](#sec-interval), or [ratio](#sec-ratio) data), we can evaluate predictions using [multiple regression](#sec-multipleRegression) or similar variants such as structural equation modeling and mixed models.

@@

To consider how we can evaluate the accuracy of predictions for a categorical outcome, consider an example adapted from @Meehl1955.
The military conducts a test of its prospective members to screen out applicants who would likely fail basic training.
To evaluate the accuracy of our predictions using the test, we can examine a confusion matrix.
A confusion matrix is a matrix that presents the predicted outcome on one dimension and the actual outcome (truth) on the other dimension.
If the predictions and outcomes are dichotomous, the confusion matrix is a 2x2 matrix with two rows and two columns that represent four possible predicted-actual combinations (decision outcomes): true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN).

@@

An example of a confusion matrix is in INSERT.

With the information in the confusion matrix, we can calculate the marginal sums and the proportion of people in each cell (in parentheses), as depicted in INSERT.

@@

Based on the marginal sums, we can compute the [marginal probabilities](#sec-baseRate), as depicted in INSERT.

@@

Based on the confusion matrix, we can calculate the prediction accuracy based on the percent accuracy of the predictions.
The percent accuracy is the number of correct predictions divided by the total number of predictions, and multiplied by 100.
In the context of a confusion matrix, this is calculated as: $100\% \times \frac{\text{TP} + \text{TN}}{N}$.
In this case, our percent accuracy was 78%—that is, 78% of our predictions were accurate, and 22% of our predictions were inaccurate.

@@

78% sounds pretty accurate.
And it is much higher than 50%, so we are doing a pretty good job, right?
Well, it is important to compare our accuracy to what accuracy we would expect to get by chance alone, if predictions were made by a random process rather than using a test's scores.
Our selection ratio was 25.4%.
How accurate would we be if we randomly selected 25.4% of people to reject?
To determine what accuracy we could get by chance alone given the selection ratio and the base rate, we can calculate the chance probability of true positives and the chance probability of true negatives.
The probability of a given cell in the confusion matrix is a [joint probability](#sec-jointProbability)—the probability of two events occurring simultaneously.
To calculate a [joint probability](#sec-jointProbability), we multiply the probability of each event.

@@

The chance expectancies for each cell of the confusion matrix are in INSERT

@@

Let us return to the confusion matrix in INSERT.
If we know the frequency of each of the four predicted-actual combinations of the confusion matrix (TP, TN, FP, FN), we can calculate sensitivity, specificity, PPV, and NPV.

>>>>

Predictions can come in different types.
Some predictions involve categorical data, whereas other predictions involve continuous data.
When dealing with a dichotomous ([nominal data](#sec-nominal) that are binary) [predictor](#sec-correlationalStudy) and [outcome](#sec-correlationalStudy) variable (or continuous data that have been dichotomized using a cutoff), we can evaluate predictions using a 2x2 table known as a [confusion matrix](#sec-confusionMatrix) (see @fig-confusionMatrix), or with logistic regression models.
When dealing with a continuous [outcome variable](#sec-correlationalStudy) (e.g., [ordinal](#sec-ordinal), [interval](#sec-interval), or [ratio](#sec-ratio) data), we can evaluate predictions using [multiple regression](#sec-multipleRegression) or similar variants such as structural equation modeling and mixed models.

@@

To consider how we can evaluate the accuracy of predictions for a categorical outcome, consider an example adapted from @Meehl1955.
The military conducts a test of its prospective members to screen out applicants who would likely fail basic training.
To evaluate the accuracy of our predictions using the test, we can examine a [confusion matrix](#sec-confusionMatrix).
A [confusion matrix](#sec-confusionMatrix) is a matrix that presents the predicted outcome on one dimension and the actual outcome (truth) on the other dimension.
If the predictions and outcomes are dichotomous, the [confusion matrix](#sec-confusionMatrix) is a 2x2 matrix with two rows and two columns that represent four possible predicted-actual combinations (decision outcomes), as in @fig-confusionMatrix: true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN).

::: {#fig-confusionMatrix}
![](images/confusionMatrix.png){fig-alt="A Two-by-Two Confusion Matrix."}

A Two-by-Two Confusion Matrix.
:::

@@

An example of a [confusion matrix](#sec-confusionMatrix) is in @fig-confusionMatrix1.

::: {#fig-confusionMatrix1}
![](images/confusionMatrix1.png){fig-alt="Example of a Two-by-Two Confusion Matrix. TP = true positives; TN = true negatives; FP = false positives; FN = false negatives."}

Example of a Two-by-Two Confusion Matrix. TP = true positives; TN = true negatives; FP = false positives; FN = false negatives.
:::

With the information in the [confusion matrix](#sec-confusionMatrix), we can calculate the marginal sums and the proportion of people in each cell (in parentheses), as depicted in @fig-confusionMatrix2.

::: {#fig-confusionMatrix2}
![](images/confusionMatrix2.png){fig-alt="Example of a Two-by-Two Confusion Matrix With Marginal Sums. TP = true positives; TN = true negatives; FP = false positives; FN = false negatives."}

Example of a Two-by-Two Confusion Matrix With Marginal Sums. TP = true positives; TN = true negatives; FP = false positives; FN = false negatives.
:::

@@

Based on the marginal sums, we can compute the [marginal probabilities](#sec-baseRate), as depicted in @fig-confusionMatrix3.

::: {#fig-confusionMatrix3}
![](images/confusionMatrix3.png){fig-alt="Example of a Two-by-Two Confusion Matrix With Marginal Sums and Marginal Probabilities. TP = true positives; TN = true negatives; FP = false positives; FN = false negatives; BR = base rate; SR = selection ratio."}

Example of a Two-by-Two Confusion Matrix With Marginal Sums and Marginal Probabilities. TP = true positives; TN = true negatives; FP = false positives; FN = false negatives; BR = base rate; SR = selection ratio.
:::

@@

Based on the [confusion matrix](#sec-confusionMatrix), we can calculate the prediction accuracy based on the percent accuracy of the predictions.
The percent accuracy is the number of correct predictions divided by the total number of predictions, and multiplied by 100.
In the context of a [confusion matrix](#sec-confusionMatrix), this is calculated as: $100\% \times \frac{\text{TP} + \text{TN}}{N}$.
In this case, our percent accuracy was 78%—that is, 78% of our predictions were accurate, and 22% of our predictions were inaccurate.

@@

78% sounds pretty accurate.
And it is much higher than 50%, so we are doing a pretty good job, right?
Well, it is important to compare our accuracy to what accuracy we would expect to get by chance alone, if predictions were made by a random process rather than using a test's scores.
Our selection ratio was 25.4%.
How accurate would we be if we randomly selected 25.4% of people to reject?
To determine what accuracy we could get by chance alone given the selection ratio and the base rate, we can calculate the chance probability of true positives and the chance probability of true negatives.
The probability of a given cell in the [confusion matrix](#sec-confusionMatrix) is a [joint probability](#sec-jointProbability)—the probability of two events occurring simultaneously.
To calculate a [joint probability](#sec-jointProbability), we multiply the probability of each event.

@@

The chance expectancies for each cell of the [confusion matrix](#sec-confusionMatrix) are in @fig-confusionMatrix4.

::: {#fig-confusionMatrix4}
![](images/confusionMatrix4.png){fig-alt="Chance Expectancies in Two-by-Two Confusion Matrix. BR = base rate; SR = selection ratio."}

Chance Expectancies in Two-by-Two Confusion Matrix. BR = base rate; SR = selection ratio.
:::

@@

Let us return to the [confusion matrix](#sec-confusionMatrix) in @fig-confusionMatrix3.
If we know the frequency of each of the four predicted-actual combinations of the [confusion matrix](#sec-confusionMatrix) (TP, TN, FP, FN), we can calculate sensitivity, specificity, PPV, and NPV.
